Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2022-08-11 writing/investigating-GPP.tex
updated paper with discussion of controls

The control stimuli in Exps.~1-3 were main clauses. For the question embedding (Exps.~1q, 2q and 3q), we used the polar questions in (2). The NRRCs in parentheses are only used in Exps.~2q and 3q to make the use of the assent diagnostics in the at-issueness blocks more natural: like the target stimuli, the control stimuli consist of two clauses, one of which the relevant speaker assents with.

\begin{exe}
\exi{(1)} Sentences for control stimuli in question embedding (Exps.~1q, 2q and 3q)
\begin{xlist}
\ex Do these muffins (, which are really delicious,) have blueberries in them?
\ex Does this pizza (, which I just made from scratch,) have mushrooms on it? 
\ex Was Jack (, who is my long-time neighbor,) playing outside with the kids? 
\ex Does Ann (, who is a local performer,) dance ballet?
\ex Were John's kids (, who are very well-behaved,) in the garage?
\ex Does Samantha (, who is really into fashion,) have a new hat?
\end{xlist}
\end{exe}

In the negation, modal and conditional embeddings, we used the positive declarative variants of (2), given in (1). Exps.~1n and 1m, we used the sentences in (1) without the NRRCs. We used the sentences in (1) with the NRRCs for Exp.~1c and for Exps.~2 and 3.

%in the projection blocks of were formed from the sentences in (1). In Exps.~1, we used the polar question variants of the sentences in (1) in Exp.~1q, and the (positive declarative) sentences in (1) in Exps.~1n, 1m, and 1c; in Exp.~1c, we added the non-restrictive relative clauses (NRRCs) provided in (1) to the respective subjects, so that the control stimuli have two clauses, like the target stimuli.

@@

\paragraph{Expectations of answers to certainty and at-issueness diagnostics} For the control stimuli in (1) used in Exps.~1q, 2q and 3q, which are polar questions, we expected participants to give low responses in the `certain that' diagnostic, indicating that the speaker is not certain of the main clause content, i.e., that the content does not project out of the question. For the control stimuli in (2), which are positive declarative sentences, we expected high responses in the `certain that' diagnostic for the control stimuli, indicating that the speaker is certain of the main clause content. These expectations were borne out, as shown in the third column of Table~\ref{t-controls}.

%In Exps.~2 and 3, the NRRCs were included in all control stimuli: the carrier sentences were the polar question variants of the sentences in (1) in Exps.~2q and 3q, and the sentences in (1) for the remaining experiments. 
%We had the same expectations as for the control stimuli in Exps.~1: low responses in the `certain that' diagnostic in Exps.~2q and 3q (indicating that the speaker is not certain of the main clause content, i.e., that the content does not project out of the question), and high responses in the `certain that' diagnostic in the remaining experiments (indicating, again, that the speaker is certain of the main clause content). These expectations were borne out, as shown in the lower nine rows of Table \ref{t-proj-controls}.

\begin{table}[h!]
\centering
\begin{tabular}{r r r r l }
Experiment & Control stimuli & Certainty &  \multicolumn{2}{l}{At-issueness} \\ & & Mean rating & Mean rating & Measure \\ 
\hline
1q & (1) & .14 & .05  & asking whether $c$ \\
2q & (1) with NRRC & .18 & .07 & {\em yes}, $c$\\
3q & (1) with NRRC & .17 &  .28 & {\em yes}, but $\neg c'$\\
\hline
1n & (2) &  .95 & .04 & sure that $c$\\
1m & (2) & .96 & .03 & sure that $c$\\
1c & (2) with NRRC & .94  & .08 & sure that $c$\\
\hline
2n & (2) with NRRC& .96 & .22 & {\em yes, that's true}, $c$\\
2m & (2) with NRRC& .96 & .25 & {\em yes, that's true}, $c$\\
2c & (2) with NRRC& .96 & .22 & {\em yes, that's true}, $c$\\
\hline
3n & (2) with NRRC & .94 & .44 & {\em yes, that's true}, but $\neg c'$\\
3m & (2) with NRRC & .93 & .50 & {\em yes, that's true}, but $\neg c'$\\
3c & (2) with NRRC & .93 & .53 & {\em yes, that's true}, but $\neg c'$\\
\hline
\end{tabular}
\caption{Mean certainty and at-issueness ratings for control stimuli, for self-declared American English participants}\label{t-controls}
\end{table}

For the at-issueness diagnostics, we expected the main clause content to be at-issue across the diagnostics. We thus expected low responses across all at-issueness diagnostics, given that participants' responses were coded so that the higher a participants' response, the more not-at-issue the content was hypothesized to be (since we investigate whether there is a positive correlation between projection and not-at-issueness). However, as shown in the fourth column of Table~\ref{t-proj-controls}, the at-issueness ratings are not always low. In particular, xx

Exp.~3q\\
A: Do these muffins, which are really delicious, have blueberries in them?\\
B: Yes, but they aren't really delicious.

Exp.~2n, 2m, 2c\\
A: These muffins, which are really delicious, have blueberries in them.\\
B: Yes, that's true, they have blueberries in them.

Exp.~3n, 3m, 3c\\
A: These muffins, which are really delicious, have blueberries in them.\\
B: Yes, that's true, but they aren't really delicious.

>>>>

The control stimuli in Exps.~1-3 were main clauses. In the question embedding experiments (Exps.~1q, 2q and 3q), the control stimuli consisted of the polar questions in (1). The NRRCs in parentheses were realized in Exps.~2q and 3q, where at-issueness was measured with an assent diagnostic: the control stimuli here consisted of two clauses (like the target stimuli), allowing the relevant speaker to assent with one of the two clauses.

\begin{exe}
\exi{(1)} Sentences for control stimuli in in question embedding experiments (Exps.~1q, 2q and 3q)
\begin{xlist}
\ex Do these muffins (, which are really delicious,) have blueberries in them?
\ex Does this pizza (, which I just made from scratch,) have mushrooms on it? 
\ex Was Jack (, who is my long-time neighbor,) playing outside with the kids? 
\ex Does Ann (, who is a local performer,) dance ballet?
\ex Were John's kids (, who are very well-behaved,) in the garage?
\ex Does Samantha (, who is really into fashion,) have a new hat?
\end{xlist}
\end{exe}

We expected participants to give low responses on the `certain that' diagnostic for the control stimuli in (1), indicating that the speaker is not certain of the main clause content, because main clause content is hypothesized to not project out of the question. These expectations were borne out, as shown in the third column of Table~\ref{t-controls} for Exps.~1q, 2q, and 3q. We also expected participants to give low responses on the at-issueness diagnostics for the control stimuli in (1), indicating that the main clause content is at-issue. These expectations were borne out for Exps.~1q and 2q, as shown in the fourth column of Table~\ref{t-controls}, but not for Exp.~3q.

In the remaining experiments, the control stimuli consisted of the positive declarative variants of (1) given in (2). The NRRCs in parentheses were realized in Exps.~2 and 3 for the reason explained above, as well as in Exp.~1c, to make the control stimuli more similar to the target stimuli (which also consisted of two clauses: the antecedent and the consequent).

@@

We expected participants to give high responses on the `certain that' diagnostic for the control stimuli in (2), indicating that the speaker is certain of the main clause content, because speakers are hypothesized to be committed to the asserted main clause content. These expectations were borne out, as shown in the third column of Table~\ref{t-controls} for the `n', `m' and `c' variants of Exps.~1, 2, and 3. We expected participants to give low responses on the at-issueness diagnostics for the control stimuli in (2), indicating that the main clause content is at-issue. These expectations were borne out for the `n', `m', and `c' variants of Exps.~1, as shown in the fourth column of Table~\ref{t-controls}, but not for the `n', `m', and `c' variants of Exps.~2 and 3.

\begin{table}[h!]
\centering
\begin{tabular}{r r r r l }
& &  \multicolumn{2}{c}{Mean ratings} &  \\ 
Exp. & Control stimuli & Certainty & At-issueness & At-issueness measure \\ 
\hline
1q & (1) & .14 & .05  & asking whether $c$ \\
1n & (2) &  .95 & .04 & sure that $c$\\
1m & (2) & .96 & .03 & sure that $c$\\
1c & (2) with NRRC & .94  & .08 & sure that $c$\\
\hline
2q & (1) with NRRC & .18 & .07 & {\em yes}, $c$\\
2n & (2) with NRRC& .96 & .22 & {\em yes, that's true}, $c$\\
2m & (2) with NRRC& .96 & .25 & {\em yes, that's true}, $c$\\
2c & (2) with NRRC& .96 & .22 & {\em yes, that's true}, $c$\\
\hline
3q & (1) with NRRC & .17 &  .28 & {\em yes}, but $\neg c'$\\
3n & (2) with NRRC & .94 & .44 & {\em yes, that's true}, but $\neg c'$\\
3m & (2) with NRRC & .93 & .50 & {\em yes, that's true}, but $\neg c'$\\
3c & (2) with NRRC & .93 & .53 & {\em yes, that's true}, but $\neg c'$\\
\hline
\end{tabular}
\caption{Mean certainty and at-issueness ratings for control stimuli, for self-declared American English participants}\label{t-controls}
\end{table}

The at-issueness measures that did not work out as expected are illustrated in the examples in (3), (4), and (5). In (3), B responds in the affirmative to A's question, but then denies the truth of the NRRC. The group mean is comparatively higher (at .28) than for the other `q' experiments, including Exp.~2q. This result suggests that B's affirmative response to A's question is taken by the participants to affirm not only the main clause content, but also the content of the NRRC. Subsequent denial of the truth of the NRRC results in degradation in acceptability.

\begin{exe}
\exi{(3)} Sample control stimulus in Exp.~3q
\begin{xlist}
\exi{A:} Do these muffins, which are really delicious, have blueberries in them?
\exi{B:} Yes, but they aren't really delicious.
\end{xlist}
\end{exe}

In (4), B assents with A's indicative assertion, and then re-affirms the truth of the main clause content. The group mean is comparatively higher (at .22 or .25) than for Exp.~2q. This result suggests that responding in the affirmative to A's question is a different speech act than assenting with A's indicative assertion: whereas it does not appear to be odd, in the latter case, to spell out the propositional antecedent of {\em yes}, it does seem to be odd to do so in the latter case.

\begin{exe}
\exi{(4)} Sample control stimulus in Exps.~2n, 2m, and 2c
\begin{xlist}
\exi{A:} These muffins, which are really delicious, have blueberries in them.
\exi{B:} Yes, that's true, they have blueberries in them.
\end{xlist}
\end{exe}

In (5), B assents with A's indicative assertion, but then denies the truth of the NRRC. Here, the group mean is quite a bit higher (at .44, .50, .53) than for Exp.~3q or for Exps.~2n, 3m, and 3c. This result suggests, again, that responding in the affirmative to A's question is a different speech act than assenting with A's indicative assertion: while denying the truth of the NRRC after responding in the affirmative to the question is already odd (Exp.~3q), denying the truth of the NRRC after assenting with A's assertion with the NRRC is worse.

\begin{exe}
\exi{(5)} Sample control stimulus in Exps.~3n, 3m, and 3c
\begin{xlist}
\exi{A:} These muffins, which are really delicious, have blueberries in them.
\exi{B:} Yes, that's true, but they aren't really delicious.
\end{xlist}
\end{exe}

The responses to the control stimuli with NRRCs that used either one of the assent diagnostics suggest that assenting with a speaker's indicative utterance that includes an NRRC includes the content of the NRRC. Two possibilities: either the content of the NRRC is already entered into the CG when A asserts the sentence (thus making reaffirmation or denial odd); or B's {\em yes, that's true} assents not just with the main clause content, but also (though perhaps a bit less?) with the content of the NRRC (thereby making reaffirmation or denial odd).
